Open Data, Open Source and Open Standards in chemistry - The Blue Obelisk five years on

### 2011-05-12 paper.tex
Added my name and contact.
Start tidying up web app section

Bioclipse in turn acts as base for other software
such as Brunn~\cite{Alvarsson:2011fk}, a laboratory information system for
microplate based high-throughput screening. Brunn provides a graphical interface
for handling different plate layouts and dilution series and can automaticly
generate dose response curves and calculate IC$_{50}$-values.

@@

While desktop software has composed the majority of scientific tools
since the computer was introduced, the internet continues to change
how applications and content are distributed and presented. As such,
there is a growing need to present chemical and scientific information
in a web-based medium. Recently, a new version of the HTML
specification, HTML5 \cite{html5}, defines a well-developed framework
for creating native web applications in JavaScript. This native
framework provides a significant benefit to scientists as the web is
an open and free medium to distribute scientific knowledge, ideas and
education. It also poses significant new problems, as the cost
associated with learning and developing tools for new technologies is
not trivial. As such, it is an important goal for the Blue Obelisk to
provide such tools and encourage the community to utilise these new
technologies to improve products and education. A specific example of
this initiative is the ChemDoodle Web Components library
\cite{ChemDoodleWeb}, produced by iChemLabs. The ChemDoodle Web
Components library is open source under the GPL v3 license with a
liberal HTML exception. ChemDoodle Web Components allow the scientist
to present publication quality 2D and 3D graphics and animations for
chemical structures, reactions and spectra. Beyond graphics, this tool
provides a framework for user interaction to create dynamic
applications through web browsers, desktop platforms and mobile
devices such as the iPhone, iPad and Android devices. It is now easier
than ever for scientists to create web-based scientific content,
without having to master complicated technologies such as Java
applets. iChemLabs is dedicated to funding, developing and supporting
the library to ensure that the next generation of scientific
applications is easily achievable by academia, government and
industry. This goal helps to make sure that the cost of education
decreases while using the web to further spread science.

Some projects have also been adopted by other fields, including biology
and crystallography. A good example
here is Jmol, which has, in the past five years, gone through two major revisions
as part of an accelerated development focus. Starting with Jmol 10.2, released in
April of 2006, Jmol 11.0 added over 100 new features and was released in March of
2007. Three years later, Jmol 12.0 followed with over 700 new features.
Based on the outstanding work of Michael (Miguel) Howard and others, this development
path took Jmol from a ``Rasmol/Chime replacement'' to a fully fledged molecular
visualisation package, including full support for crystallography~\cite{Hanson2010},
efficient delivery of surface data using fast solvent-excluded surface
generation and the highly compressed JVXL format [http://chemapps.stolaf.edu/jmol/docs/misc/JVXL-format.pdf],
 display of molecular orbitals from standard basis set/coefficient data,
the inclusion of dynamic minimisation using the UFF force field, and
a full implementation of Daylight SMILES[ref] and SMARTS[ref], with extensions to
conformational and biomolecular substructure searching in the form of Jmol
BioSMARTS.[http://jmol.svn.sourceforge.net/viewvc/jmol/trunk/Jmol/src/org/jmol/smiles/package.html]
Behind the scenes, Jmol has been essentially completely rewritten. A full
mathematical expression evaluator was added, and the scripting language
was folded into a versatile JavaScript-like command language, including a
complete set of program flow commands, several new data types, and
user-defined variables and functions. An exciting non-chemical application
of Jmol has been its incorporation into the open source Sage
Mathematics Project [http://www.sagemath.org/], where Jmol is being used
to deliver three-dimensional depiction of mathematical functions.

>>>>

Bioclipse in turn acts as base for other software
such as Brunn~\cite{WebBrunn}, a laboratory information system for
microplate based high-throughput screening. Brunn provides a graphical interface
for handling different plate layouts and dilution series and can automaticly
generate dose response curves and calculate IC$_{50}$-values.

@@

While desktop software has composed the majority of scientific tools
since the computer was introduced, the internet continues to change
how applications and content are distributed and presented. The web
presents new opportunities for scientists as it
an open and free medium to distribute scientific knowledge, ideas and
education. Web applications are software that runs within the browser,
typically implemented in Java or JavaScript.
Recently, a new version of the HTML
specification, HTML5, defines a well-developed framework
for creating native web applications in JavaScript and opens up
new possibilities for visualising chemical data.

Jmol, the interactive 3D molecular viewer, is one of the most widely used
chemistry applets, and indeed has
seen widespread use in other fields such as biology and 
even mathematics (it is used for 3D depiction of mathematical
functions in the Sage Mathematics Projects~\cite{WebSage}). It is implemented
in Java, and has gone from being a ``Rasmol/Chime'' replacement to a fully fledged molecular
visualisation package, including full support for crystallography~\cite{Hanson2010},
display of molecular orbitals from standard basis set/coefficient data,
the inclusion of dynamic minimisation using the UFF force field, and
a full implementation of Daylight SMILES and SMARTS, with extensions to
conformational and biomolecular substructure searching (Jmol
BioSMARTS).

A specific example of
this initiative is the ChemDoodle Web Components library
\cite{ChemDoodleWeb}, produced by iChemLabs. The ChemDoodle Web
Components library is open source under the GPL v3 license with a
liberal HTML exception. ChemDoodle Web Components allow the scientist
to present publication quality 2D and 3D graphics and animations for
chemical structures, reactions and spectra. Beyond graphics, this tool
provides a framework for user interaction to create dynamic
applications through web browsers, desktop platforms and mobile
devices such as the iPhone, iPad and Android devices. It is now easier
than ever for scientists to create web-based scientific content,
without having to master complicated technologies such as Java
applets. iChemLabs is dedicated to funding, developing and supporting
the library to ensure that the next generation of scientific
applications is easily achievable by academia, government and
industry. This goal helps to make sure that the cost of education
decreases while using the web to further spread science.
